---
title: "Getting started with vibewidget"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with vibewidget}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What this is

`vibewidget` generates interactive, LLM-authored notebook widgets from a
plain-English description and a `data.frame`, and renders them inline in
Positron, RStudio, Quarto, and R Markdown - with no JavaScript written by
you. It is a thin R client of the `vibe_widget` Python engine, run
in-process through `reticulate`; there is no Jupyter kernel anywhere in the
stack.

```{r}
library(vibewidget)

w <- vw_create(
  "temperature trends by region with a year slider",
  weather_df,
  outputs = list(selection = "selected rows")
)
w
```

The first time any `vw_*` function actually touches Python, `reticulate`
provisions a small Python environment (via `uv`) with the engine's
dependencies - this is a one-time cost per machine, cached afterward.
`vw_create()` returns immediately; the printed widget streams generation
progress live.

## Reading state back into R

```{r}
w$status()             # "generating" | "ready" | "error" | "blocked"
w$wait()                # block until a terminal status, if you need to
w$outputs$selection()   # current value - NULL until something is selected
```

To react to changes as they happen, use `vw_observe()`. Because R is
single-threaded, this cannot be a live push from a background thread - it
polls (via the `later` package) only when R is otherwise idle at the
console, so it never blocks or interrupts your work:

```{r}
vw_observe(w, "selection", function(v) message(nrow(v), " rows selected"))
```

## Pushing state into the widget, and calling actions

```{r}
w$inputs$threshold <- 25
w$actions$reset()
```

## Revising a widget in English

```{r}
vw_edit(w, "make the legend a dropdown")
```

For structural or configuration changes rather than element-targeted
revision, `w$edit(...)` is equivalent.

## Saving and loading

```{r}
vw_save(w, "trends.vw")
w2 <- vw_load("trends.vw", data = weather_df)

vw_html(w, "trends.html")  # a fully static, self-contained export
```

## Configuration

```{r}
vw_config(model = "google/gemini-3.1-pro-preview")
vw_options()
```

## Rendered Quarto/R Markdown output

There is no R process behind a rendered `.qmd`/`.Rmd` document, so
`vw_create()` automatically falls back to a fully static, self-contained
snapshot when it detects a knitr render (or any non-interactive session) -
the widget still renders, but its state does not round-trip back into R.
Live, two-way widgets belong in the Positron/RStudio console or a notebook
cell; a rendered document gets whichever state existed when it was
rendered. Force one mode or the other with `options(vibewidget.mode =
"live")`/`"static"` if the automatic detection ever guesses wrong.
